---
title: "Prediction Assignment Writeup"
author: "Henry lopez (hlopezo@uni.pe)"
date: "12 de agosto de 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Summary
This document is the final report of the Peer Assessment project from the Practical Machine Learning course, which is a part of the Data Science Specialization. It was written and coded in RStudio, using its knitr functions and published in the html format. The purpose of this analysis is to predict the manner in which the six participants performed the exercises described below and to answer the questions of the associated course quiz. The machine learning algorithm, which uses the classe variable in the training set, is applied to the 20 test cases available in the test data. The predictions are submitted to the Course Project Prediction Quiz for grading. 
# 2. Background
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

# 3. Data Loading and Exploratory Analysis
# 3.1 Dataset Overview
The training data for this project are available here:

  *   https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv
   
The test data are available here:

  *   https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv


# 3.2 Environment Preparation
We first upload the R libraries that are necessary for the complete analysis.
```{r}
rm(list=ls())  # free up memory for the download of the data sets
```
# 3.3 Data downloading

Data downloading testing
```{r}
if (!file.exists("pml-testing.csv")) {
  url.testing <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
  download.file(url.testing, destfile = "pml-testing.csv")
}
```

Data downloading training
```{r}
if (!file.exists("pml-training.csv")) {
  url.training <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
  download.file(url.training, destfile = "pml-training.csv")
}
```
Asign the data to variables
```{r}
data_train <- read.csv('pml-training.csv', strip.white = TRUE, na.strings = c("NA",""))
data_quiz  <- read.csv('pml-testing.csv',  strip.white = TRUE, na.strings = c("NA",""))
```

```{r}
dim(data_train)
dim(data_quiz)
```
Load the library
```{r}
#load
library(tibble)
library(bitops)
library(lattice)
library(ggplot2)
library(caret)
library(rpart)
library(rpart.plot)
library(corrplot)
library(rattle)
library(randomForest)
library(RColorBrewer)
```
Create two partitions (75 % and 25 %) within the original training dataset.
```{r}
in_train  <- createDataPartition(data_train$classe, p=0.75, list=FALSE)
train_set <- data_train[ in_train, ]
test_set  <- data_train[-in_train, ]


dim(train_set)
dim(test_set)
```


The two datasets (train_set and test_set) have a large number of NA values as well as near-zero-variance (NZV) variables. Both will be removed together with their ID variables.
```{r}
nzv_var <- nearZeroVar(train_set)
train_set <- train_set[ , -nzv_var]
test_set  <- test_set [ , -nzv_var]
dim(train_set)
dim(test_set)
```

Remove variables that are mostly NA. A threshlod of 95 % is selected.

```{r}
na_var <- sapply(train_set, function(x) mean(is.na(x))) > 0.95
train_set <- train_set[ , na_var == FALSE]
test_set  <- test_set [ , na_var == FALSE]
dim(train_set)
dim(test_set)

```
Since columns 1 to 5 are identification variables only, they will be removed as well.
```{r}
train_set <- train_set[ , -(1:5)]
test_set  <- test_set [ , -(1:5)]
dim(train_set)
dim(test_set)
```

# 4 . Correlation Analysis

Perform a correlation analysis between the variables before the modeling work itself is done. Select "FPC" for the first principal component order.
```{r}
set.seed(12082020)
corr_matrix <- cor(train_set[ , -54])
                corrplot(corr_matrix, 
                         order = "FPC", 
                         method = "circle", 
                         type = "lower",
                           tl.cex = 0.6,
                         tl.col = rgb(0, 0, 0))

```

If two variables are highly correlated their colors are either dark blue (for a positive correlation) or dark red (for a negative corraltions). To further reduce the number of variables, a Principal Components Analysis (PCA) could be performed as the next step. However, since there are only very few strong correlations among the input variables, the PCA will not be performed. Instead, a few different prediction models will be built next.

                
# 5. Prediction Models
            
## 5.1. Decision Tree Model
```{r}
set.seed(12082020)
fit_decision_tree <- rpart(classe ~ ., data = train_set, method="class")
fancyRpartPlot(fit_decision_tree)

```



Predictions of the decision tree model on test_set.

```{r}
set.seed(12082020)
predict_decision_tree <- predict(fit_decision_tree, newdata = test_set, type="class")
conf_matrix_decision_tree <- confusionMatrix(predict_decision_tree, test_set$classe)
conf_matrix_decision_tree
```




Plot the predictive accuracy of the decision tree model.
```{r}
set.seed(12082020)
plot(conf_matrix_decision_tree$table, 
     col = conf_matrix_decision_tree$byClass, 
     main = paste("Decision Tree Model: Predictive Accuracy =",
                  round(conf_matrix_decision_tree$overall['Accuracy'], 4)))

```

## 5.2. Generalized Boosted Model (GBM)


```{r}

set.seed(12082020)
ctrl_GBM <- trainControl(method = "repeatedcv", number = 5, repeats = 2)
fit_GBM  <- train(classe ~ ., data = train_set,
                  method = "gbm",
                  trControl = ctrl_GBM, 
                  verbose = FALSE)
fit_GBM$finalModel

```
Predictions of the GBM on test_set.
```{r}
set.seed(12082020)

predict_GBM <- predict(fit_GBM, newdata = test_set)
conf_matrix_GBM <- confusionMatrix(predict_GBM, test_set$classe)
conf_matrix_GBM

```


## 5.3. Random Forest Model

```{r}
set.seed(12082020)
ctrl_RF <- trainControl(method = "repeatedcv", number = 5, repeats = 2)
fit_RF  <- train(classe ~ ., data = train_set, 
                 method = "rf",
                 trControl = ctrl_RF, 
                 verbose = FALSE)
fit_RF$finalModel
```

Predictions of the Random Forest model on test_set.

```{r}
set.seed(12082020)
predict_RF <- predict(fit_RF, newdata = test_set)
conf_matrix_RF <- confusionMatrix(predict_RF, test_set$classe)
conf_matrix_RF

```





# 6. Applying the Best Predictive Model to the Test Data


The Random Forest model is selected and applied to make predictions on the 20 data points from the original testing dataset (data_quiz).

```{r}
predict_quiz <- predict(fit_RF, newdata = data_quiz)
predict_quiz
```

